Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-03 Multiresolution_framework/ltx/manuscript.tex
overal edits
fixes figure captions

\subsection{Experiment I}
The spatial frequency response of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. This is shown in Fig.~\ref{fig:ObservationFrequencyResponce} where the graph obtained by averaging the spectral power of the spatial frequency (over time) from the observations \citep{Scerri2009}. The spatial bandwidths of wavelets was calculated analytically using equation \eqref{eq:MRA-Wavelettransform}. The result suggested that wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field which yields $n_x = 131$ states, corresponding to 9 scaling functions at $j=0$ and 8, 16, 32 and 66 wavelet functions, respectively at $j=0$, $j=1$, $j=2$ and $j=3$. 
\begin{figure}[!t]\label{fig:ObservationFrequencyResponce}
 	\centering
 		\includegraphics[scale=1]{./Graph/Figure2.pdf}
 		\caption{{\bf Spatial frequency analysis}. The average (over time) power in dB of the spatial frequency of the observations.}
 \end{figure}

The EM algorithm was allowed to run until a maximum number of 20 iterations was reached, though typically the change in the transition matrix Frobenius norm dropped below $10^{-6}$ after less than 10 iterations. The rate of convergence of the EM based algorithm is depicted in Fig.~\ref{fig:MRA-Convergence}. 
\begin{figure}[!t]\label{fig:MRA-Convergence} 
 	\centering
 		\includegraphics[scale=1]{./Graph/ConvergenceRate.pdf}
 		\caption{{\bf Convergence of the EM algorithm}. Representative plot of
the $\mathbf A(\boldsymbol\theta)$ Frobenius norm vs iterations of EM based algorithm. The change in the stopping criterion falls below $10^{-6}$ after less than 10 iterations.}
 \end{figure}     
% The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, obtained by averaging the spectral power of the spatial frequency (over time) from the observations \cite{Scerri2009}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

The actual connectivity kernel and the decomposition is plotted in Fig.~\ref{fig:KernelEstimate}(A). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Fig.~\ref{fig:KernelEstimate}(B) and (C) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions was used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$ accordingly. 
\begin{figure}[!t]
\includegraphics{./Graph/Figure5.pdf}
\includegraphics{./Graph/Figure6.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{a}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{b}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}

% \begin{figure}[!h] 
% 	\centering
% 		\includegraphics[scale=1]{./Graph/fig2.pdf}
% 		\caption{Connectivity kernel estimate. (a) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (b)-(c) Kernel scaling and wavelets functions, respectively, at $j=1$ used in estimation algorithm.}
% 	\label{fig:KernelEstimate}
% \end{figure}
 One snap shot of the field reconstruction, at different levels of approximation is given in Fig.~\ref{fig:FieldEstimates100}, showing small discrepancy between the actual and estimated field at $j=3$. From this figure, it can be observed that the overall trend of the underlying field can be also captured at the coarser levels.
\begin{figure}[!t]
\centering
\includegraphics{./Graph/Figure4a.pdf}\\
\includegraphics{./Graph/Figure4b.pdf}\\
\includegraphics{./Graph/Figure4c.pdf}\\
\includegraphics{./Graph/Figure4d.pdf}\\
\caption{ {\bf Examples of the actual and estimated neural fields for experiment I}. Actual (grey) and estimated (black) neural fields at a single time instant for different spatial resolutions. (\textbf{a}) $j=0$, $n_x=17$. (\textbf{b}) $j=1$, $n_x=33$. (\textbf{c}) $j=2$, $n_x=65$. (\textbf{d}) $j=3$, $n_x=131$.}
% (\textbf{e}) $j=4$, $n_x=263$.
\label{fig:FieldEstimates100}
\end{figure}
\subsection{Experiment II}
To study the performance of different models with various MRA capabilities, a single realization was obtained using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} under the same circumstances of the previous experiment. The generated data set was then employed to estimate the underlying field and the connectivity kernel parameters using different state-space models, $j=0$ up to $j=4$. The RMSE of the field estimation for different models is shown in Fig.~\ref{fig:RMSE}. By increasing the value of $j$, error in the estimation decreases and converges to a steady value of 0.71, confirming that the model with $j=3$ is in fact adequate to capture the dynamics of the underlying field. The spatiotemporal characteristics of the reconstructed fields (A-E) along with the actual field (F)  are shown in Fig.~\ref{fig:FieldEstimation}. The experiment was performed several times and the results were consistent over each run.
 \begin{figure}[!t]	\label{fig:RMSE} 
 	\centering
 		\includegraphics[scale=1]{./Graph/Figure3.pdf}
 		\caption{{\bf Error in the field reconstruction for experiment II}. The mean (over time) of average (over space) RMSE of the estimated field of state-space models with different MRA capabilities. $j=0$, $RMSE = 1.83$; $j=1$, $RMSE = 1.41$; $j=2$, $RMSE = 0.83$; $j=3$, $RMSE = 0.71$; $j=4$, $RMSE=0.71$}
 \end{figure}

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top+\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top.
\end{align}
Rearranging  equation \eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top.
\end{equation}
Substituting \ref{eq:Xi0derivation1} into \ref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t$.
%\begin{equation}\label{eq:Xi1derivation1}
% \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
%\end{equation}
%substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
%\begin{equation}
 %\boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
%\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names} 
\bibliography{MRAIDE}

>>>>

\subsection{Experiment I}
The spatial frequency response of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. This is shown in Fig.~\ref{fig:ObservationFrequencyResponce} where the graph obtained by averaging the spectral power of the spatial frequency (over time) from the observations \citep{Scerri2009}. The spatial bandwidths of wavelets was calculated analytically using equation \eqref{eq:MRA-Wavelettransform}. The result suggested that wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field which yields $n_x = 131$ states, corresponding to 9 scaling functions at $j=0$ and 8, 16, 32 and 66 wavelet functions, respectively at $j=0$, $j=1$, $j=2$ and $j=3$. 
\begin{figure}[!t]
 	\centering
 		\includegraphics[scale=1]{./Graph/Figure2.pdf}
 		\caption{{\bf Spatial frequency analysis}. The average (over time) power in dB of the spatial frequency of the observations.}  
\label{fig:ObservationFrequencyResponce} 
 \end{figure}

The EM algorithm was allowed to run until a maximum number of 20 iterations was reached, though typically the change in the transition matrix Frobenius norm dropped below $10^{-6}$ after less than 10 iterations. The rate of convergence of the EM based algorithm is depicted in Fig.~\ref{fig:MRA-Convergence}. 
\begin{figure}[t]
 	\centering
 		\includegraphics[scale=1]{./Graph/ConvergenceRate.pdf}
 		\caption{{\bf Convergence of the EM algorithm}. Representative plot of
the $\mathbf A(\boldsymbol\theta)$ Frobenius norm vs iterations of EM based algorithm. The change in the stopping criterion falls below $10^{-6}$ after less than 10 iterations.} 
\label{fig:MRA-Convergence}  
 \end{figure}     
% The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, obtained by averaging the spectral power of the spatial frequency (over time) from the observations \cite{Scerri2009}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

The actual connectivity kernel and the decomposition is plotted in Fig.~\ref{fig:KernelEstimate}(A). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Fig.~\ref{fig:KernelEstimate}(B) and (C) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions was used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$ accordingly. 
\begin{figure}[!t]
\includegraphics{./Graph/Figure5.pdf}
\includegraphics{./Graph/Figure6.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{A}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{B}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}

% \begin{figure}[!h] 
% 	\centering
% 		\includegraphics[scale=1]{./Graph/fig2.pdf}
% 		\caption{Connectivity kernel estimate. (a) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (b)-(c) Kernel scaling and wavelets functions, respectively, at $j=1$ used in estimation algorithm.}
% 	\label{fig:KernelEstimate}
% \end{figure}
 One snap shot of the field reconstruction, at different levels of approximation is given in Fig.~\ref{fig:FieldEstimates100}, showing small discrepancy between the actual and estimated field at $j=3$. From this figure, it can be observed that the overall trend of the underlying field can be also captured at the coarser levels.
\begin{figure}[!t]
\centering
\includegraphics{./Graph/Figure4a.pdf}\\
\includegraphics{./Graph/Figure4b.pdf}\\
\includegraphics{./Graph/Figure4c.pdf}\\
\includegraphics{./Graph/Figure4d.pdf}\\
\caption{ {\bf Examples of the actual and estimated neural fields for experiment I}. Actual (grey) and estimated (black) neural fields at a single time instant for different spatial resolutions. (\textbf{a}) $j=0$, $n_x=17$. (\textbf{b}) $j=1$, $n_x=33$. (\textbf{c}) $j=2$, $n_x=65$. (\textbf{d}) $j=3$, $n_x=131$.}
% (\textbf{e}) $j=4$, $n_x=263$.
\label{fig:FieldEstimates100}
\end{figure}
\subsection{Experiment II}
To study the performance of different models with various MRA capabilities, a single realization was obtained using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} under the same circumstances of the previous experiment. The generated data set was then employed to estimate the underlying field and the connectivity kernel parameters using different state-space models, $j=0$ up to $j=4$. The RMSE of the field estimation for different models is shown in Fig.~\ref{fig:RMSE}. By increasing the value of $j$, error in the estimation decreases and converges to a steady value of 0.71, confirming that the model with $j=3$ is in fact adequate to capture the dynamics of the underlying field. The spatiotemporal characteristics of the reconstructed fields (A-E) along with the actual field (F)  are shown in Fig.~\ref{fig:FieldEstimation}. The experiment was performed several times and the results were consistent over each run.
 \begin{figure}[!t]	 
 	\centering
 		\includegraphics[scale=1]{./Graph/Figure3.pdf}
 		\caption{{\bf Error in the field reconstruction for experiment II}. The mean (over time) of average (over space) RMSE of the estimated field of state-space models with different MRA capabilities. $j=0$, $RMSE = 1.83$; $j=1$, $RMSE = 1.41$; $j=2$, $RMSE = 0.83$; $j=3$, $RMSE = 0.71$; $j=4$, $RMSE=0.71$}
\label{fig:RMSE}
 \end{figure}

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top+\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top.
\end{align}
Rearranging  equation \eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top.
\end{equation}
Substituting \ref{eq:Xi0derivation1} into \ref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t$.
%\begin{equation}\label{eq:Xi1derivation1}
% \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
%\end{equation}
%substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
%\begin{equation}
 %\boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
%\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names}  
%\bibliographystyle{plain}

\bibliography{MRAIDE}
